PLANETESIMAL ACCRETION AT SHORT ORBITAL PERIODS

### 2022-08-02 main.tex
Add reference to childs 22
Rework intro, fill in citations

% TRQ: the following paragraph (particularly the last sentence could
% use citations: probably to review articles.
Planetesimal accretion is one of a number of stages in which micron-sized solids from the protostellar nebula coalesce to eventually build terrestrial planets. In the earliest stages, aerodynamic forces dominate the growth and evolution of the solids. Millimeter-sized bodies grow through adhesive pairwise collisions and stay well-coupled to the surrounding gas. Beyond this size, however, a number of growth barriers present themselves. Most notably, larger solids orbit the central star at Keplerian speeds as they decouple from the gas, which orbits at a sub-keplerian speed due to radial pressure support. This leads the solids to feel a headwind, which is maximally effective at sapping away angular momentum for objects around 1 meter in size. At this size, the timescale for the growing solids to fall onto the star is catastrophically short and leads to what is known as the drift barrier. In addition, two-body collisions between mm- to cm-sized bodies tend to result in bouncing or destruction, rather than continued growth. For these reasons, a number of mechanisms have been proposed which facilitate fast growth from mm to km sizes by locally concentrating solids. Dust traps, streaming instability, pebble piles...etc.

Beyond kilometer scales, gravity begins to dominate and aerodynamic gas drag plays a smaller and smaller role. During this phase, collision cross sections are enhanced as gravitational focusing \citep{safronov69} acts to bend the trajectories of bodies undergoing close encounters. Large bodies are most effective at focusing the trajectories of nearby planetesimals, leading to a period of runaway growth \citep{wetherill89, kokubo96, barnes09}. Eventually, the largest bodies (known as oligarchs) dynamically heat the surrounding planetesimals, severely limiting further growth \citep{kokubo98}. The end result of this phase is a bimodal population of dynamically cold oligarchs, surrounded by dynamically hot, difficult to accrete residual planetesimals. Lines of evidence suggest that the asteroid belt, kuiper belt and the oort cloud are largely composed of the leftovers of this stage of planet formation. (Mention more specific evidence, morbidelli 09 paper, CAIs?)

Although gas drag has a minimal influence on the Moon to Mars-sized oligarchs, it is enough to prevent these largest bodies from perturbing each other onto crossing orbits. Simulations show that evaporation of the gas disk is required to allow instability to trigger a phase of giant impacts (Mention that disk fraction decay timescale roughly matches timing of giant impacts in SS). It is during this phase that oligarchs collide to form Earth-sized planets \citep{chambers98, raymond06}.

Over the last few decades, terrestrial planet formation models have largely advanced by matching properties of the solar system. Compared to exoplanetary systems, the solar system provides a rich set of constraints (isotopic ratios, cratering records, small body populations) that are mostly unmeasurable for even the closest neighboring planet forming systems. However, the system architectures discovered by spaced-based missions in the last decade reveal that the solar system could very well be an outlier in terms of what a typical planet-forming disk produces. In addition, the sizes and compositions of the terrestrial planets likely rely on a series of finely-tuned events to play out that involve truncation of the primordial planetesimal disk \citep{raymond17}, inward, followed by outward migration of an outer giant planet \citep{walsh11}, or a large-scale instability triggered by a pair of convergently migrating giant planets \citep{tsiganis05, levison11, nesvorny11}. Given qualitatively similar initial conditions, solar system formation models can even occasionally reproduce the correct masses and orbital periods of the terrestrial planets without invoking any of the aforementioned scenarios, given the right random number seed \citep{fischer14}.

Given the difficult question of whether to treat the solar system as
an outlier, the best way forward is to use statistical samples of
exoplanetary architectures to develop and inform formation
models. This is generally done through the use of population synthesis
models \citep{ida04, alibert11}, but many of the mechanisms in these
models are informed and tuned by solar system constraints. One
pervasive and exotic result revealed by the Kepler space telescope is the discovery of hundreds of compact multi-planet terrestrial systems, dubbed systems of tightly-packed inner planets (STIPs). Although there is no formal definition of a STIP, they typically contain 3 or more Earth-sized planets with orbital periods extending between 1 and 100 days. Reconciling the structural differences between the solar system (devoid of large bodies interior to 88 days) and STIPs is going to be an important step in building a general, widely-applicable planet formation model.

To date, a large body of work exists that has attempted to reproduce
the architectures of STIPs, starting from planetary embryos (cite some
examples). However, the runaway and oligarchic growth phases, which
precede the assembly of the embryos, are assumed to be
ubiquitous. Given that the timescales for accretion and gravitational
scattering scale differently with encounter velocity, which itself
scales with orbital period, it is not entirely clear that planetary
embryos should form in the same way close to the star as they do at
much longer, more thoroughly studied orbital periods. In this paper,
we use direct N-body simulations to explore the outcome of the
planetesimal accretion stage at orbital periods shorter than 100 days
around an M star.
% TRQ perhaps mention habitable zone here.  Somewhere, a discussion of
% the importance of understanding habitable zones around M-stars,
% where STIPS are found, needs to be discussed.
In particular, we seek to understand what the orbital and mass distributions of the embryos and residual planetesimals look like, and to assess whether the initial conditions used by late-stage simulations of STIP assembly are reasonable.

@@

>>>>

Planetesimal accretion is a key phase in the terrestrial planet growth process, bridging the gap from kilometer-sized bodies up to roughly moon-sized objects known as planetary embryos. In the earliest stages of the planet formation process, aerodynamic forces dominate the growth and evolution of the solids and statistical models \citep{johansen14, birnstiel16} are appropriate to describe how these numerous, small bodies coagulate. Due to the internal pressure support of the gas disk, the gas itself orbits at sub-Keplerian speed and exerts a headwind on any solids large enough to decouple from the gas \citep{weidenschilling77}. Above about a meter in size, this headwind is maximally effective at sapping away angular momentum and planet-building material can fall onto the central star on catastrophically short timescales \citep{weidenschilling77, nakagawa86}. Additionally, laboratory experiments suggest that collisions between mm- to cm- sized solids tend to result in bouncing or destruction, rather than continued growth \citep{blum93, beitz11, colwell03}. For these reasons, a number of mechanisms which involve radially concentrating solids in a planet-forming disk have been proposed to facilitate fast growth from mm to km sizes \citep{johansen07, lyra08, bai10}. Interestingly, formation models for the short-period multiplanet systems revealed by Kepler \citep{fabrycky14} also seem to require enhanced concentrations of planet-building material to reproduce the observed architectures \citep{raymond07, hansen12}.

Regardless of how the mm- to km-sized growth barriers are surmounted, gravity begins to dominate and aerodynamic gas drag plays a less significant role beyond this size. During this phase, collision cross sections are enhanced as gravitational focusing \citep{safronov69} acts to bend the trajectories of bodies undergoing close encounters. Because larger bodies are more effective at focusing the trajectories of nearby planetesimals, a period of runaway growth occurs \citep{wetherill89, kokubo96, barnes09}. Eventually, the largest bodies (known as oligarchs) dynamically heat the surrounding planetesimals, severely limiting further growth \citep{kokubo98}. The end result of this phase is a bimodal population of dynamically cold oligarchs, surrounded by dynamically hot, difficult to accrete residual planetesimals. Lines of evidence suggest that the asteroid belt, Kuiper belt and the Oort cloud are largely composed of the leftovers of this stage of planet formation.

After a long period of quiescence, the collection of embryos and remaining planetesimals undergoes a large-scale instability \citep{chambers98}. Suddenly, the oligarchics are no longer on isolated, stable orbits and coalesce to form Earth-sized planets through a series of extremely energetic, giant impacts \citep{kokubo02, raymond05, raymond06}. Due to the relative ease of modeling the early dust coagulation phases and the final giant impact phase, these steps in the terrestrial planet formation process have received the most attention in the literature. The planetesimal accretion phase, which we will focus on in this paper, falls into an awkward in-between, where there are too many particles to directly track with traditional N-body codes, while the gravitational influence of the few, but massive oligarchs that form make statistical methods an inappropriate approach. Because of this limitation, planetesimal accretion is usually modeled in a narrow ring and the results are then scaled to suit whatever situation is being studied. N-body simulations of terrestrial planet formation typically begin with a series of neatly spaced oligarchs, whose mass varies smoothly with orbital period. As we will show in this paper, this extrapolation with orbital period is inappropriate, particularly at short ($<$ 10 day) periods.

Given that systems of tightly-packed inner planets (STIPs) appear to be a common outcome of planet formation, understanding exactly how solids accumulate at short orbital periods is crucial. Although gas-disk driven migration of the planets themselves is often invoked to explain the observed architectures, we will focus on an in-situ model in this paper. That is, once the planetesimals themselves form, they largely stay in place and any subsequent large-scale movement of the solids are the result of mutual gravitational interactions. The focus of this work will be to understand how the outcome of the planetesimal accretion process scales with orbital period by using a high-powered N-body code to directly follow the growth and evolution of the planetesimals across a wide range of orbital periods (1 to 100 days). In doing so, we will assess whether the typical `isolation mass' initial conditions used in studies of terrestrial planet formation are actually appropriate for understanding STIPs. In a series of follow up papers, we plan to use these results to directly track the formation of full-sized planets from planetesimals and to understand how gravitational perturbations from a nearby gas giant might affect the outcome of this process.

% This stuff doesn't fit since I reworked the intro. Might be better to save some of this for paper 2 (planetesimals->planets)
% Still deciding whether I should mention habitability and low mass stars
% I think there are plenty of STIPs around sun like stars, but I should double check this

%Over the last few decades, terrestrial planet formation models have largely advanced by matching properties of the solar system. Compared to exoplanetary systems, the solar system provides a rich set of constraints (isotopic ratios, cratering records, small body populations) that are mostly unmeasurable for even the closest neighboring planet forming systems. However, the system architectures discovered by spaced-based missions in the last decade reveal that the solar system could very well be an outlier in terms of what a typical planet-forming disk produces. In addition, the sizes and compositions of the terrestrial planets likely rely on a series of finely-tuned events to play out that involve truncation of the primordial planetesimal disk \citep{raymond17}, inward, followed by outward migration of an outer giant planet \citep{walsh11}, or a large-scale instability triggered by a pair of convergently migrating giant planets \citep{tsiganis05, levison11, nesvorny11}. Given qualitatively similar initial conditions, solar system formation models can even occasionally reproduce the correct masses and orbital periods of the terrestrial planets without invoking any of the aforementioned scenarios, given the right random number seed \citep{fischer14}.

%Given the difficult question of whether to treat the solar system as
%an outlier, the best way forward is to use statistical samples of
%exoplanetary architectures to develop and inform formation
%models. This is generally done through the use of population synthesis
%models \citep{ida04, alibert11}, but many of the mechanisms in these
%models are informed and tuned by solar system constraints. One
%pervasive and exotic result revealed by the Kepler space telescope is the discovery of hundreds of compact multi-planet terrestrial systems, dubbed systems of tightly-packed inner planets (STIPs). Although there is no formal definition of a STIP, they typically contain 3 or more Earth-sized planets with orbital periods extending between 1 and 100 days. Reconciling the structural differences between the solar system (devoid of large bodies interior to 88 days) and STIPs is going to be an important step in building a general, widely-applicable planet formation model.

%To date, a large body of work exists that has attempted to reproduce
%the architectures of STIPs, starting from planetary embryos (cite some
%examples). However, the runaway and oligarchic growth phases, which
%precede the assembly of the embryos, are assumed to be
%ubiquitous. Given that the timescales for accretion and gravitational
%scattering scale differently with encounter velocity, which itself
%scales with orbital period, it is not entirely clear that planetary
%embryos should form in the same way close to the star as they do at
%much longer, more thoroughly studied orbital periods. In this paper,
%we use direct N-body simulations to explore the outcome of the
%planetesimal accretion stage at orbital periods shorter than 100 days
%around an M star.
% TRQ perhaps mention habitable zone here.  Somewhere, a discussion of
% the importance of understanding habitable zones around M-stars,
% where STIPS are found, needs to be discussed.
%In particular, we seek to understand what the orbital and mass distributions of the embryos and residual planetesimals look like, and to assess whether the initial conditions used by late-stage simulations of STIP assembly are reasonable.
%

@@

Radial mixing appears to be enhanced by a partial collision model, along with smaller collision cross section inflation \citep{childs22}.
